Let's talk about Thurstone & Co.: An information-theoretical model for comparative judgments, and its statistical translation

### 2024-11-18 index.qmd
section:issues; process:write

<!-- 1. Overview of CJ method in assessing traits -->
In *comparative judgment* (CJ) studies, judges assess a specific trait or attribute across various stimuli by performing pairwise comparisons [@Thurstone_1927a; @Thurstone_1927b]. Each comparison produces a dichotomous outcome, indicating which stimulus is perceived to exhibit a higher trait level. For example, when assessing text quality, judges compare pairs of written texts (the stimuli) to determine the relative quality each text exhibit (the trait) [@Laming_2004; @Pollitt_2012b; @Whitehouse_2012; @vanDaal_et_al_2016; @Lesterhuis_2018; @Coertjens_et_al_2017; @Goossens_et_al_2018; @Bouwer_et_al_2023].

@@

<!-- 2. What is the effectiveness of CJ -->
Numerous studies have documented the effectiveness of CJ in assessing traits and competencies over the past decade. These studies have emphasized three aspects of the method's effectiveness: its reliability, validity, and practical applicability. Research on reliability indicates that CJ requires a relatively small number of pairwise comparisons [@Verhavert_et_al_2019; @Crompvoets_et_al_2022] to produce trait scores that are as precise and consistent as those generated by other assessment methods [@Coertjens_et_al_2017; @Goossens_et_al_2018; @Bouwer_et_al_2023]. Furthermore, evidence suggests that the reliability and time efficiency of CJ are comparable, if not superior, to those of other assessment methods when employing adaptive comparison algorithms [@Pollitt_2012b; @Verhavert_et_al_2022; @Mikhailiuk_et_al_2021]. On the other hand, research on validity suggests that scores generated by CJ can accurately represent the traits under measurement [@Whitehouse_2012; @vanDaal_et_al_2016; @Lesterhuis_2018; @Bartholomew_et_al_2018; @Bouwer_et_al_2023]. Finally, research on practical applicability highlights the method's versatility across both educational and non-educational contexts [@Kimbell_2012; @Jones_et_al_2015; @Bartholomew_et_al_2018; @Jones_et_al_2019; @Marshall_et_al_2020; @Bartholomew_et_al_2020; @Boonen_et_al_2020].

@@

<!-- 1. Thurstone's Theory of CJ: discriminal process -->
In its most general form, Thurstone's theory [-@Thurstone_1927b] posits that the dichotomous outcome resulting from comparing two stimuli is determined by two factors: the discriminal process of each stimulus and their discriminal difference. The *discriminal process* refers to the psychological effect each stimulus has on the judges, or more simply stated, the judges’ perception of the trait level of each stimulus. Thurstone assumes that the discriminal process for each stimulus follows a Normal distribution. In this distribution, the mode (mean), known as the *modal discriminal process*, represents the position of the stimulus on the trait continuum, while the dispersion, known as the *discriminal dispersion*, reflects variability in the stimulus' perceived trait level. @fig-discriminal_process shows example distributions of discriminal process for two stimuli (objects).

@@

<!-- 2. But what are limitations of Case V (and the BTL model) -->
However, Case V (and its BTL model counterpart) was originally developed to provide a "rather coarse scaling" of traits [@Thurstone_1927b, pp. 269], prioritizing statistical simplicity over precision in trait measurement [@Kelly_et_al_2022, pp. 677]. As a result, its assumptions may not suit applications beyond the psychophysical contexts for which it was created. Thurstone himself cautioned that its use "should not be made without (an) experimental test" [@Thurstone_1927b, pp. 270], acknowledging that some assumptions could prove problematic with more complex, less homogeneous stimuli, such as handwriting or English compositions [@Thurstone_1927a, pp. 374]. Consequently, given that current CJ applications often deal with complex, less homogeneous stimuli, two key assumptions of Case V may not consistently hold in theory or practice: the zero correlation and equal dispersion between stimuli.

@@

<!-- 3. The problems of assuming zero correlation between stimuli -->
[The *zero correlation between stimuli* represents two additional assumptions about the method:   ]{style="color:red;"}

<!-- this translates into the cancellation of judges effects by mean of opposing and equally weighted 'mood' and 'simultaneous contrasts' effects. This is demonstrated using the additive nature of the logit scale, which helps to cancel 'bias' judges effects. -->

@@

<!-- The accuracy of judgment may vary by subject and by judge [@Gill_et_al_2013; @vanDaal_2020; @vanDaal_et_al_2017] -->

<!-- If it is acknowledged that judges may differ in what they value, it is not inconceivable that judges may differ as a function of their gender, age, ethnicity, culture, class, income, education and so on. Would the aggregate view of young, British-Asian men always be the same as the aggregate view of older, black women? [@Kelly_et_al_2022, pp. 683] -->

@@

<!-- this also translates into the idea that stimuli are the main focus of estimation and analysis, but what happens when the focus of analysis is the individuals that generated those stimuli. Meaning there is an amount of correlation that it is not accounted for. example of Boonen -->

@@

<!--   use the example of writing!!! -->

@@

@@

<!-- 1. The BTL model as a measurement model for estimating latent traits -->
Building on the previous section, it is evident that the BTL model typically  functions as the measurement model for the trait of interest [@Andrich_1978; @Bramley_2008]. A measurement model specifies how manifest variables contribute to the estimation of latent variables [@Everitt_et_al_2010]. For example, when evaluating text quality, researchers use the BTL model to process the dichotomous outcomes resulting from the pairwise comparisons (the manifest variables) to estimate scores that reflect the underlying texts' quality level (the latent variable) [@Laming_2004; @Pollitt_2012b; @Whitehouse_2012; @vanDaal_et_al_2016; @Lesterhuis_2018; @Coertjens_et_al_2017; @Goossens_et_al_2018; @Bouwer_et_al_2023].

@@

<!-- []{style="color:red;"}. -->

>>>>

<!-- 1. Overview of CJ method in assessing traits -->
In *comparative judgment* (CJ) studies, judges assess a specific trait or attribute across various stimuli by performing pairwise comparisons [@Thurstone_1927a; @Thurstone_1927b]. Each comparison produces a dichotomous outcome, indicating which stimulus is perceived to exhibit a higher trait level. For example, when assessing text quality, judges compare pairs of written texts (the stimuli) to determine the relative quality each text exhibit (the trait) [@Laming_2004; @Pollitt_2012b; @Whitehouse_2012; @vanDaal_et_al_2016; @Lesterhuis_2018_thesis; @Coertjens_et_al_2017; @Goossens_et_al_2018; @Bouwer_et_al_2023].

@@

<!-- 2. What is the effectiveness of CJ -->
Numerous studies have documented the effectiveness of CJ in assessing traits and competencies over the past decade. These studies have emphasized three aspects of the method's effectiveness: its reliability, validity, and practical applicability. Research on reliability indicates that CJ requires a relatively small number of pairwise comparisons [@Verhavert_et_al_2019; @Crompvoets_et_al_2022] to produce trait scores that are as precise and consistent as those generated by other assessment methods [@Coertjens_et_al_2017; @Goossens_et_al_2018; @Bouwer_et_al_2023]. Furthermore, evidence suggests that the reliability and time efficiency of CJ are comparable, if not superior, to those of other assessment methods when employing adaptive comparison algorithms [@Pollitt_2012b; @Verhavert_et_al_2022; @Mikhailiuk_et_al_2021]. On the other hand, research on validity suggests that scores generated by CJ can accurately represent the traits under measurement [@Whitehouse_2012; @vanDaal_et_al_2016; @Lesterhuis_2018_thesis; @Bartholomew_et_al_2018; @Bouwer_et_al_2023]. Finally, research on practical applicability highlights the method's versatility across both educational and non-educational contexts [@Kimbell_2012; @Jones_et_al_2015; @Bartholomew_et_al_2018; @Jones_et_al_2019; @Marshall_et_al_2020; @Bartholomew_et_al_2020; @Boonen_et_al_2020].

@@

<!-- 1. Thurstone's Theory of CJ: discriminal process -->
In its most general form, Thurstone's theory [-@Thurstone_1927b] posits that the dichotomous outcome resulting from comparing two stimuli is determined by two factors: the discriminal process of each stimulus and their discriminal difference. The *discriminal process* refers to the psychological effect each stimulus has on the judges, or more simply stated, the judges’ perception of the trait level of each stimulus. Thurstone assumes that the discriminal process for each stimulus follows a Normal distribution. In this distribution, the mode (mean), known as the *modal discriminal process*, represents the position of the stimulus on the trait continuum, while the dispersion, known as the *discriminal dispersion*, reflects the variability in the stimulus' perceived trait level. @fig-discriminal_process shows example distributions of discriminal process for two stimuli (objects).

@@

<!-- 2. But what are limitations of Case V (and the BTL model) -->
However, Case V (and its BTL model counterpart) was originally developed to provide a "rather coarse scaling" of traits [@Thurstone_1927b, pp. 269], prioritizing statistical simplicity over precision in trait measurement [@Kelly_et_al_2022, pp. 677]. As a result, its assumptions may not suit applications beyond the psycho-physical contexts for which it was created. Thurstone himself cautioned that its use "should not be made without (an) experimental test" [@Thurstone_1927b, pp. 270], acknowledging that some assumptions could prove problematic with more complex, less homogeneous stimuli, such as handwriting or English compositions [@Thurstone_1927a, pp. 374]. Consequently, given that current CJ applications often deal with complex, less homogeneous stimuli, two key assumptions of Case V may not consistently hold in theory or practice: the zero correlation and equal dispersion between stimuli.

@@

<!-- 3. What does the zero correlation between stimuli imply? -->
The assumption of *zero correlation between stimuli* is best illustrated with an example. For instance, when evaluating text quality, this assumption suggests that a judge’s perception of quality in one text does not influence their perception of the same trait in the comparison text. Thurstone attributes this independence to the cancellation of potential judges' biases, driven by opposing and equally weighted factors; commonly known as 'mood' and 'simultaneous contrast' effects, that operate during pairwise comparisons [@Thurstone_1927b, pp. 268]. This cancellation of bias has been mathematically demonstrated by @Andrich_1978, assuming additive biases and a logit scale derived from the BTL model.

<!-- 4. But what are the problems of assuming zero correlation? -->
However, given that evidence has shown judges' quality assessments are influenced by multiple, complex aspects of the texts [@vanDaal_et_al_2016; @Lesterhuis_2018; @Chambers_et_al_2022], it is plausible that the accuracy of these judgments also varies based on factors such as age, culture, education, expertise [@Kelly_et_al_2022, pp. 683], and ultimately, based on the individual judge themselves [@Gill_et_al_2013; @vanDaal_et_al_2017; @vanDaal_2020].

[This means that the cancellation of judges' biases may not hold in theory or practice, and that judges' biases may depend on the stimuli, by means of an interaction (but present) or because the judges works as a confounder ]{style="color:red;"}

Such differences may not be detected through analyses of bias and misfit [@Kelly_et_al_2022, pp. 683].

<!-- Would the aggregate view of young, British-Asian men always be the same as the aggregate view of older, black women? [@Kelly_et_al_2022, pp. 683] -->

<!-- The accuracy of judgment may vary by subject and by judge [@Gill_et_al_2013; @vanDaal_et_al_2017; @vanDaal_2020] -->

@@

@@

<!-- this also translates into the idea that stimuli are the main focus of estimation and analysis, but what happens when the focus of analysis is the individuals that generated those stimuli. Meaning there is an amount of correlation that it is not accounted for. Use example of Boonen!! -->

@@

@@

<!-- Of these, the most questionable is the one that the discriminal dispersions of all the objects are equal. Thurstone clearly did not expect this to hold for any but the most simple stimuli. The scripts used in cross-moderation exercises are obviously far more complex than any used by Thurstone, so it seems naïve to expect this assumption to hold here [@Bramley_2008]. -->

@@

<!-- 1. The BTL model as a measurement model for estimating latent traits -->
Building on the previous section, it is evident that the BTL model commonly functions as the measurement model for the trait of interest in CJ experiments [@Andrich_1978; @Bramley_2008]. A measurement model specifies how manifest variables contribute to the estimation of latent variables [@Everitt_et_al_2010]. For example, when evaluating text quality, researchers use the BTL model to process the dichotomous outcomes resulting from the pairwise comparisons (the manifest variables) to estimate scores that reflect the underlying quality level of texts (the latent variable) [@Laming_2004; @Pollitt_2012b; @Whitehouse_2012; @vanDaal_et_al_2016; @Lesterhuis_2018_thesis; @Coertjens_et_al_2017; @Goossens_et_al_2018; @Bouwer_et_al_2023].

@@

<!-- []{style="color:red;"} -->
